### 2025-05-16 docs/article/Sections/state-of-the-art.tex
chapter 2 - include text about Web Servers approaches and Async APIs Support.

\section{State of the Art}

In this section, we first present the main properties that characterize each web
template technology approach, along with the advantages and drawbacks resulting
from these characteristics.

Next, we delve into the different design choices adopted by web servers in their
internal architectures, and how these choices impact the behavior of the web
template engines described in the previous subsection.

\subsection{Related Work}

@@

@@

\subsubsection{Asynchronous support}

@@

1. Nem se quer Existe API para iterar sobre um Async data model. 
   E.g JStachio, .... apenas suportam interface Iterable. => Compilation Error.
2. Consegue usar qq API e.g Kotlinx.Html and Groovy  => usa mal => produzir HTML out of order
3. Sim suporta => well-formed HTML => Thymleaf e o HtmlFlow

>>>>

\section{Background and Related Work}

In this section, we first examine the different design choices adopted by web 
servers in their internal architectures, and how these choices impact the 
behavior of the web template engines.
Then, in Subsection~2.2, we present the main properties that characterize each 
web template technology approach, along with the advantages and drawbacks that 
result from these characteristics.

In traditional thread-per-request architectures, each incoming request is
handled by a dedicated thread. The web server maintains a thread pool from
which a thread is assigned to handle each request. As the load increases, the
number of active threads can grow rapidly, potentially exhausting the pool.
This can lead to performance and scalability issues, as the system may become
bogged down by context switching and thread management
overhead~\cite{kant2000scalable}. Spring MVC is one of the most widely used
frameworks that follows this model.

On the other hand, in modern \textit{low-thread} architectures, the server uses
a small number of threads to handle a large number of requests.
\textit{Low-thread} servers, also known as
\textit{event-driven}~\cite{event-driven-servers}, offer a significant
advantage in efficiently managing a high number of concurrent I/O operations
with minimal resource usage.

The \textit{non-blocking} I/O model employed in low-thread servers is
well-suited for handling large volumes of data asynchronously~\cite{Meijer12}.
This combination of low-thread servers and asynchronous data models has
facilitated the development of highly scalable, responsive, and resilient web
applications capable of managing substantial data loads~\cite{Jin15}.
The prominence of this concept increased with the advent of Node.js in
2009, and subsequently, various technologies adopted this approach
in the Java ecosystem, including Netty, Akka HTTP, Vert.X, and Spring WebFlux.
% Among these, Spring WebFlux stands out as the most widely used web framework in
% Java, according to surveys like JetBrains' State of Java report (2021) and
% community metrics from platforms like Github and StackOverflow.
The \textit{non-blocking} I/O model in low-thread servers functions optimally
only when HTTP handlers avoid blocking.
Therefore, HTML templates need to be proficient in dealing with the asynchronous
APIs provided by data models.
While most legacy web templates struggle with asynchronous models, DSLs for HTML
face no such limitations, leveraging all constructions available in the host
programming language.
However, the unexpected intertwining of asynchronous handlers' completion and
HTML builders' execution may potentially lead to malformed HTML with an
unexpected layout, as demonstrated in the subsequent subsection~\ref{sec:async-support}.

An alternative approach involves utilizing user-level threads, while maintaining
a blocking I/O and a synchronous programming paradigm.
However, this approach still requires a user-level I/O subsystem capable of
mitigating system-level blocking, which is crucial for the performance of
I/O-intensive applications. This technique offers a lightweight solution for
efficiently managing a larger number of concurrent sessions by minimizing
per-thread overhead.
In 2020, Karsten~\cite{karsten2020} demonstrated how this strategy supports a
synchronous programming style, effectively abstracting away the complexities
associated with managing continuations in asynchronous programming.
A similar idea has been followed in Kotlin with
\textit{coroutines}~\cite{elizarov2021coroutines}, and most recently in the Java
standard library with \textit{virtual threads} released in Java 21.
% Both Spring WebFlux and Vert.X web frameworks provide native support for Kotlin
% \textit{coroutines}, enabling developers to utilize a specific
% \texttt{Dispatcher} to switch the execution thread context to a suitable
% subsystem and reduce \textit{system-level} blocking.
However, no work has yet leveraged this mechanism to enable legacy template
engines to provide PSSR while preserving non-blocking progress.
% We will demonstrate this capability by comparing it with our proposal of
% \textit{suspendable web templates} in a benchmark using Spring WebFlux.

% Determining the appropriate size of \emph{N} for the I/O susbsystem may hinge on
% factors such as the nature of the workload and the frequency of
% \textit{Blocking} I/O operations.

\subsection{Web Templates}

@@

Before detailing each of the aforementioned characteristics, Table~11 presents 
a breakdown of state-of-the-art template engines, classified according to the 
identified properties.

@@

\subsubsection{Asynchronous support}
\label{sec:async-support}

@@

1. Nem se quer Existe API para iterar sobre um Async data model. 
   E.g JStachio, .... apenas suportam interface Iterable. => Compilation Error.

2. Consegue usar qq API e.g Kotlinx.Html and Groovy  => usa mal => produzir HTML out of order

3. Sim suporta => well-formed HTML => Thymleaf e o HtmlFlow

One of the reasons for legacy web templates not supporting asynchronous APIs
is the absence of a unified standard calling convention for asynchronous calls.
While there is a single, straightforward way to use a synchronous API with a
direct style, where the result of a method call corresponds to its returned
value, there is no equivalent standard in the asynchronous approach. 
Instead, we may encounter various asynchronous conventions depending on the
programming language and runtime environment.
Some of these approaches include \emph{continuation-passing style}
(CPS)~\cite{scheme}, \textit{promises}~\cite{promise} async/await
idiom~\cite{async_await}, reactive streams \cite{ReactiveStreams}, Kotlin
Flow~\cite{kotlinlang}, and others.

Many \textit{general-purpose languages} (GPLs) have embraced the
\texttt{async}/\texttt{await} feature~\cite{async_await} enabling non-blocking
routines to mimic the structure of synchronous ones, allowing developers to
reason about instruction flow sequentially.
The simplicity and broad adoption of this programming model have led to its
incorporation into mainstream languages like C\#, JavaScript, Python, Perl,
Swift, Kotlin, and others, excluding Java.
However, implementing \texttt{async}/\texttt{a}wait requires compiler support to
translate \textit{suspension points} (i.e., \texttt{await} statements) into state
machines. Most template engines operate using an external DSL with their own
templating dialect (e.g., Thymeleaf, JSP, Jade, Handlebars, and others), which do not
inherently leverage asynchronous capabilities from their host GPLs.
